\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-07-15 papers/jss/article.tex
some changes to the jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@305 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{

	\pkg{depmixS4} implements a general framework for defining and estimating
	 dependent mixture models in the \proglang{R} programming language
	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
	models, and latent class and finite mixture distribution models. The models can
	be fitted on mixed multivariate data with distributions from the \code{glm} 
	family, the logistic multinomial, or the multivariate normal distribution. Other 
	distributions can be added easily, and an example is provided. Parameter 
	estimation is done through an EM algorithm or by a direct optimization approach using
	the \pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
	Parameters can be estimated subject to general linear constraints.  
}

@@

\Plainkeywords{hidden markov model, dependent mixture model, mixture model} %% without formatting
%% at least one keyword must be supplied

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, latent Markov models are commonly used as regime
switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models. See e.g.\ \citet[][chapter~1]{Cappe2005}
for an overview of different applications.

The \pkg{depmixS4} package was motivated by the fact that Markov
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models.  Common programs for
Markovian models include Panmark \citep{Pol1996}, and for latent class
models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
number of important features, besides not being freely available.
There are currently some packages in R that handle hidden Markov
models but they lack a number of features that we needed in our
research.  In particular, \pkg{depmixS4} was designed to meet the
following goals:

@@

	\item to be able to handle parameter estimates subject to general
	linear (in)equality constraints

@@

	\item to allow for easy extensibility, in particular, to be able
	to add new response distributions, both univariate and
	multivariate, and similarly to be able to allow for the addition
	of other transition models, e.g., continuous time observation
	models

@@

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$.  In those cases, there are no time dependencies between
observed data, and the model reduces to a finite mixture model, or a
latent class model.  Although there are other specialized packages to
deal with mixture data, one specific feature that we needed ourselves
which is to the best of our knowledge not available in other packages
is the possibility to include covariates on the prior probabilities of
class membership and the transition probabilities. In the next section, 
an outline is provided of the model and the likelihood equations.

@@

The data considered here, has the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment.  The data consists of three
variables, response time, accuracy and a covariate which is a pay-off
determining the relative reward for speeded and/or accuracy responding. These
variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Pol1996} are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series 
\citep[][, chapter~1]{Cappe2005}.  
In the next sections, the
likelihood and estimation procedures for the dependent mixture model is
described for data of the above form.  These models are called
dependent mixture models because one of the authors (Ingmar Visser)
thought it was time for a new name for these models\footnote{Only
later did I find out that \citet{Leroux1992} already coined the term
dependent mixture models in an application with hidden Markov mixtures
of Poisson count data.}.

@@

	\item a set $\vc{S}$ of latent classes or states $S_{i},\, i=1,
	\ldots , n$,

	\item matrices $\mat{A}$ of transition probabilities $a_{ij}$ for
	the transition from state $S_{i}$ to state $S_{j}$,

	\item a set $\vc{B}$ of observation density functions $b_j^k(\cdot)$ that
	provide the conditional densities of observations $O_{t}^k$ 
	associated with latent state $S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$
\end{enumerate}

The dependent mixture model is defined by the following equations: 
\begin{align}
	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
	O_{t} &= b(O_{t}|S_{t}), 
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $b$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $\vc{B}$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log likelihood can be written as
\begin{equation}
\log \Prob(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log \Prob(S_1|\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $S_t$. In the Expectation step,
we replace these with their expected values given a set of (initial) parameters 
$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
and observations $O_{1:T}$. The expected log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
\ln \Prob(O_t|S_t=j,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the r.h.s. of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{nnet}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.

@@

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

@@

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{verbatim}
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
	family=list(gaussian(),multinomial()),
	 transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod)
\end{verbatim}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{verbatim}
	> summary(fm)
	...
	Response model(s) for state 1

@@

	Response model for response 2 
	Model of type multinomial, formula: corr ~ 1
	Coefficients: 
		 [,1]     [,2]
	[1,]    0 2.245514
	Probalities at zero values of the covariates.
	0.09573715 0.9042629 	
\end{verbatim}
As can be seen, state 1 has fast response times around chance level, 
whereas state 2 corresponds with slower responding at higher accuracy 
levels.

@@

Similarly to the transition matrix, covariates on the prior 
probabilities of the latent states (or classes in this case), are 
defined by using a one-sided formula: 
\begin{verbatim}
balance$age <- balance$age-5
set.seed(1)
mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
        family=list(multinomial(), multinomial(), multinomial(),
		multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
		prior=~age, initdata=balance)
fm <- fit(mod)
\end{verbatim}
Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{verbatim}
> summary(fm)
Mixture probabilities model 
Model of type multinomial, formula: ~age
		[,1]   [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
...
\end{verbatim}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

To see the ordering of parameters use the following:
\begin{verbatim}
mod <- setpars(mod, value=1:npar(mod))
mod
\end{verbatim}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
\begin{verbatim}
mod <- setpars(mod, getpars(mod,which="fixed"))
mod
\end{verbatim}

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
\begin{verbatim}
trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	nstates=2,family=list(gaussian(),multinomial()),
	trstart=trst,inst=c(.999,0.001))
fm1 <- fit(mod)
\end{verbatim}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix (parameters numbers~6 and~10):
\begin{verbatim}
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm,equal=conpat)
\end{verbatim}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

In the \code{speed} data example, it may be more appropriate to model
the response times as an exgaus distribution rather than using the
gaussian.  We can do so as follows, first we define our own
exgaus-class extending the response-class:
\begin{verbatim}
	setClass("exgaus", contains="response")
\end{verbatim}

@@

The constructor function is defined as: 
\begin{verbatim}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
	standardGeneric("exgaus"))
setMethod("exgaus", 
	signature(y="ANY"), 
	function(y,pstart=NULL,fixed=NULL, ...) {
		y <- matrix(y,length(y))
		x <- matrix(1)
		parameters <- list()
		npar <- 3
		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
		if(!is.null(pstart)) {
			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
			parameters$mu <- pstart[1]
			parameters$sigma <- log(pstart[2])
			parameters$nu <- log(pstart[3])
		}
		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
		mod
	}
)
\end{verbatim}

The \code{fit} function is defined as follows: 
\begin{verbatim}
setMethod("fit","exgaus",
	function(object,w) {
		if(missing(w)) w <- NULL
		y <- object@y
		fit <- gamlss(y~1,weights=w,family=exGAUS(),
			control=gamlss.control(n.cyc=100,trace=FALSE),
			mu.start=object@parameters$mu,
			sigma.start=exp(object@parameters$sigma),
			nu.start=exp(object@parameters$nu))
		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
		object <- setpars(object,pars)
		object
	}
)
\end{verbatim}

@@

We first create all the response models that we want as a double list: 
\begin{verbatim}
rModels <- list(
	list(
		exgaus(rt,pstart=c(5,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
	),
	list(
		exgaus(rt,pstart=c(6,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
	)
)
\end{verbatim}
Next, we define the transition and prior probability models using the 
transInit function (which produces a transInit model, which also extends 
the response class): 
\begin{verbatim}
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
\end{verbatim}
Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
\begin{verbatim}
mod <- makeDepmix(response=rModels,transition=transition,
prior=inMod,stat=FALSE)
fm <- fit(mod)
\end{verbatim}

Using \code{summary} will print the fitted parameters. Note that the 
use of \code{makeDepmix} allows the possibility of, say, fitting a 
gaussian in one state and an exgaus distribution in another state.

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
for a large variety of response distributions. It can also fit latent class regression
and finite mixture models, although it should be noted that more specialized packages 
are available for doing so, e.g.\ \pkg{FlexMix} \cite{Leisch2004}. The package is intended
for modeling of (individual) time series data with the aim of characterizing the transition 
processes underlying the data. The possibility to use covariates on the transition matrix
greatly enhances the flexibility of the model. The EM algorithm uses a very general 
interface that allows easy addition of new response models.

>>>>

%% an abstract and keywords
\Abstract{
	\pkg{depmixS4} implements a general framework for defining and estimating
	 dependent mixture models in the \proglang{R} programming language
	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
	models, and latent class and finite mixture distribution models. The models 
	can be fitted on mixed multivariate data with distributions from the 
	\code{glm} family, the logistic multinomial, or the multivariate normal 
	distribution. Other distributions can be added easily, and an example is 
	provided. Parameter are estimated by the EM algorithm or, when (linear) 
	constraints are imposed on the parameters, by direct numerical optimization 
	with the \pkg{Rdonlp2} routine.
}

@@

\Plainkeywords{hidden Markov model, dependent mixture model, mixture model} %% without formatting
%% at least one keyword must be supplied

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, common latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and \citealp{Ghysels1994}).
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models. For more examples of applications, 
see e.g.\ \citet[][chapter~1]{Cappe2005}.

The \pkg{depmixS4} package was motivated by the fact that while Markov
models are used commonly in the social sciences, no comprehensive
package was available for fitting such models.  Common software for 
estimating Markovian models include Panmark \citep{Pol1996}, and for latent class
models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
number of important features, besides not being freely available.
There are currently some packages in \proglang{R} that handle hidden Markov
models but they lack a number of features that we needed in our
research. In particular, \pkg{depmixS4} was designed to meet the
following goals:

@@

	\item to be able to estimate parameters subject to general
	linear (in)equality constraints

@@

	\item to be easily extensible, in particular, to allow users to
  easily add new uni- or multivariate response distributions, and similarly 
  add of other transition models, e.g., continuous time observation
	models.

@@

Although \pkg{depmixS4} was designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case when
$T=1$.  In this case, there are no time dependencies between
observed data and the model reduces to a finite mixture or
latent class model. Although there are other specialized packages to
deal with mixture data, as far as we know these 
don't allow the inclusion of covariates on the prior probabilities of
class membership. The possibility to estimating the effects of covariates on 
prior and transition probabilities is a distinguishing feature of 
\pkg{depmixS4}. In the next section, an outline is provided of the model and 
the likelihood equations.

@@

The data considered here have the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment. The data consists of three
variables, response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accuracy responding. 
These variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.  However, 
common estimation procedures, such as those implemented in
\citet{Pol1996}, are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series 
\citep[][, chapter~1]{Cappe2005}.  
In the next sections, the
likelihood equation and estimation procedures for the dependent mixture model 
are described for data of the above form. We use the term
``dependent mixture model'' because one of the authors (Ingmar Visser)
thought it was time for a new name for these models\footnote{Only
later did I find out that \citet{Leroux1992} already coined the term
dependent mixture models in an application with hidden Markov mixtures
of Poisson count data.}.

@@

	\item a set $\mathcal{S}$ of latent classes or states $S_{i},\, i=1,
	\ldots , n$,

	%\item a set of matrices $\mat{A}_t$ of transition probabilities $a_{ijt} = P_t(S_j|S_i)$ 
	%for the transition from state $S_{i}$ to state $S_{j}$ at time $t$,

	\item a set $\mathcal{A}$ of transition probability functions $a_{ij}(\vc{x}_t) = 
	Pr(S_t=j|S_{t-1}=i,\vc{x}_t)$, giving the probability of a transition from state 
	$i$ to state $j$ with covariate $\vc{x}_t$,

	\item a set $\mathcal{B}$ of observation density functions $b_j^k(\vc{x}_t) = p(O_{t}^k|S_t = j, \vc{x}_t)$ that
	provide the conditional densities of observations $O_{t}^k$ 
	associated with latent class/state $j$ and covariate $\vc{x}_t$,

	\item a set $\mathcal{P}$ of latent class/state initial probability functions
	$\pi_{i}(\vc{x}_1) = Pr(S_1 = i|\vc{x}_1)$, giving the probability of class/state
	$i$ at time $t=1$ with covariate $\vc{x}_1$.

The dependent mixture model is defined by the following equations: 
%\begin{align}
%	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
%	O_{t} &= b(O_{t}|S_{t}), 
%\end{align}
\begin{align}
	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
\end{align}
%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
%density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $\vc{B}$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log likelihood can be written as
\begin{equation}
\log \Prob(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log \Prob(S_1|\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $S_t$. In the Expectation step,
we replace these with their expected values given a set of (initial) parameters 
$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
and observations $O_{1:T}$. The expected log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\ln \Prob(O^k_t|S_t=j,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the r.h.s. of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{MASS}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.

@@

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

@@

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the Gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
	  family=list(gaussian(),multinomial()),
	  transition=~scale(Pacc),instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
...
Response model(s) for state 1

@@

Response model for response 2 
Model of type multinomial, formula: corr ~ 1
Coefficients: 
	 [,1]     [,2]
[1,]    0 2.245514
Probalities at zero values of the covariates.
0.09573715 0.9042629 	
\end{CodeOutput}
\end{CodeChunk}
As can be seen, state 1 has fast response times around chance level, 
whereas state 2 corresponds with slower responding at higher accuracy 
levels.

Note that by specifying multivariate observations in terms of a list, the 
variables are considered conditionally independent (given the states). 
Conditionally \emph{dependent} variables must be handled as a single element in
the list. Effectively, this means specifying a multivariate response model. The 
only multivariate response model currently implemented in \pkg{depmix} is for
multivariate normal variables.

@@

Similarly to the transition matrix, covariates on the prior 
probabilities of the latent states (or classes in this case), are 
defined by using a one-sided formula: 
\begin{CodeChunk}
\begin{CodeInput}
> balance$age <- balance$age-5
> set.seed(1)
> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
    family=list(multinomial(), multinomial(), multinomial(),
    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
    prior=~age, initdata=balance)
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Mixture probabilities model 
Model of type multinomial, formula: ~age
   [,1]   [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
...
\end{CodeOutput}
\end{CodeChunk}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

To see the ordering of parameters use the following:
\begin{CodeChunk}
\begin{CodeInput}
> mod <- setpars(mod, value=1:npar(mod))
> mod
\end{CodeInput}
\end{CodeChunk}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
\begin{CodeChunk}
\begin{CodeInput}
> mod <- setpars(mod, getpars(mod,which="fixed"))
> mod
\end{CodeInput}
\end{CodeChunk}

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
\begin{CodeChunk}
\begin{CodeInput}
> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial()),
	  trstart=trst,inst=c(.999,0.001))
> fm1 <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix (parameters numbers~6 and~10):
\begin{CodeChunk}
\begin{CodeInput}
# start with fixed and free parameters
> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
> conpat[6] <- conpat[10] <- 2
> fm2 <- fit(fm,equal=conpat)
\end{CodeInput}
\end{CodeChunk}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

In the \code{speed} data example, it may be more appropriate to model
the response times as an exgaus distribution rather than using the
Gaussian.  We can do so as follows, first we define our own
exgaus-class extending the response-class:
\begin{CodeChunk}
	setClass("exgaus", contains="response")
\end{CodeChunk}

@@

The constructor function is defined as: 
\begin{CodeChunk}
\begin{CodeInput}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
	standardGeneric("exgaus"))
setMethod("exgaus", 
	signature(y="ANY"), 
	function(y,pstart=NULL,fixed=NULL, ...) {
		y <- matrix(y,length(y))
		x <- matrix(1)
		parameters <- list()
		npar <- 3
		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
		if(!is.null(pstart)) {
			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
			parameters$mu <- pstart[1]
			parameters$sigma <- log(pstart[2])
			parameters$nu <- log(pstart[3])
		}
		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
		mod
	}
)
\end{CodeInput}
\end{CodeChunk}

The \code{fit} function is defined as follows: 
\begin{CodeChunk}
\begin{CodeInput}
setMethod("fit","exgaus",
	function(object,w) {
		if(missing(w)) w <- NULL
		y <- object@y
		fit <- gamlss(y~1,weights=w,family=exGAUS(),
			control=gamlss.control(n.cyc=100,trace=FALSE),
			mu.start=object@parameters$mu,
			sigma.start=exp(object@parameters$sigma),
			nu.start=exp(object@parameters$nu))
		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
		object <- setpars(object,pars)
		object
	}
)
\end{CodeInput}
\end{CodeChunk}

@@

We first create all the response models that we want as a double list: 
\begin{CodeChunk}
\begin{CodeInput}
rModels <- list(
	list(
		exgaus(rt,pstart=c(5,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
	),
	list(
		exgaus(rt,pstart=c(6,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
	)
)
\end{CodeInput}
\end{CodeChunk}
Next, we define the transition and prior probability models using the 
transInit function (which produces a transInit model, which also extends 
the response class): 
\begin{CodeChunk}
\begin{CodeInput}
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
\end{CodeInput}
\end{CodeChunk}
Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
\begin{CodeChunk}
\begin{CodeInput}
mod <- makeDepmix(response=rModels,transition=transition,
prior=inMod,stat=FALSE)
fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

Using \code{summary} will print the fitted parameters. Note that the 
use of \code{makeDepmix} allows the possibility of, say, fitting a 
Gaussian in one state and an exgaus distribution in another state.

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
for a large variety of response distributions. It can also fit latent class regression
and finite mixture models, although it should be noted that more specialized packages 
are available for doing so, e.g.\ \pkg{FlexMix} \citep{Leisch2004}. The package is intended
for modeling of (individual) time series data with the aim of characterizing the transition 
processes underlying the data. The possibility to use covariates on the transition matrix
greatly enhances the flexibility of the model. The EM algorithm uses a very general 
interface that allows easy addition of new response models.
